taxadb: A High-Performance Local Taxonomic Database Interface

### 2020-01-23 paper/manuscript.Rmd
Minor typos/adjustments

As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names. How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008, @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

- Queries require internet access at all times
- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name
- The type of query is highly limited by the API design. For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?"
- Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection).  Many names providers update the server data either continuously or at regular intervals, including both revising existing names and adding new names.  <!-- For instance, @predicts2014 observes that 2013 Catalogue of Life includes the genus "Notiophilus" in two separate families, contrary to the rules of nomenclature.  This error does not occur in the 2018 edition. -->

@@

`taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.  `taxadb` will work with most DBI-compliant databases such as MySQL or Postgres [see @DBI], but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.  These latter options are also much easier for most users, since each can be installed directly as an R package. `taxadb` will default to the fastest available option.  `taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the large names providers.

@@

Taxonomic identifiers provide a fundamental abstraction which lies at the heart of managing taxonomic names.  For instance, by resolving scientific names to identifiers, we can idetnify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized. Unmatched names may indicate an error in data entry or otherwise warrant further investigation. Taxon identifiers are also easily resolved to the original authority (scientific publication) establishing the name. (The historical practice of appending an author and year to a scientific name, e.g. *Poa annua ssp. annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010]).

@@

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care (Language from vignette). Spell check of input names is outside the scope of `taxadb`.

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009, @Franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

As well getting a detailed look at specific species or ids. For example, why can't `get_ids` resolve a seemingly common species?

```{r}
## example of sci name matching to more than one id.
filter_name("Zosterops rendovae") %>% 
  select(1:5) %>%
  printtable(caption = "Accepted name matches for Zosterops rendovae.")

>>>>

As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names. How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned datasets.
These challenges have become particularly acute as it becomes increasingly common for researchers to work across a larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise for all clades needed to resolve scientific names [@Patterson2010].

While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
Here, we present an R package, `taxadb`, which seeks to address this gap.

@@

- Queries require internet access at all times.
- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name.
- The type of query is highly limited by the API design. For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?".
- Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection).  Many names providers update the server data either continuously or at regular intervals, including both revising existing names and adding new names.

@@

`taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.  `taxadb` will work with most DBI-compliant databases such as MySQL or Postgres, but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.  These latter options are also much easier for most users, since each can be installed directly as an R package. `taxadb` will default to the fastest available option.  `taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the large names providers.

@@

Taxonomic identifiers provide a fundamental abstraction which lies at the heart of managing taxonomic names.  For instance, by resolving scientific names to identifiers, we can identify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized. Unmatched names may indicate an error in data entry or otherwise warrant further investigation. Taxon identifiers are also easily resolved to the original authority (scientific publication) establishing the name. (The historical practice of appending an author and year to a scientific name, e.g. *Poa annua ssp. annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010]).

@@

`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons. First, a name may match to multiple accepted names, as in the case of a species that has been split. By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information. A name may also be unresolved due to typos or improper formatting. `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g. `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name. These modifications are not appropriate in all settings and should be used with care. Spell check of input names is outside the scope of `taxadb`.

Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym. Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009; @Franz2018]. The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".

@@

As well as giving a detailed look at specific species or ids. For example, why can't `get_ids` resolve a seemingly common species?

```{r}
multi_match <- filter_name("Zosterops rendovae")
```

```{r echo = FALSE, cache = FALSE}
multi_match %>%
  select(1:5) %>%
  printtable(caption = "Accepted name matches for Zosterops rendovae.")
```
